---
title: "Vignette"
author: "Rachel Fellman"
date: "2023-10-02"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This document is intended to explain how to contact and API and query, parse, and return readable data from said API. I will be accessing the earthquake api from the USGS.

# Required Packages
I used the following packages to create the vignette:
  * `httr`: for accessing the API
  * `jsonlite`: to convert API data into a data frame
  * `tidyverse`: used for piping, plotting, and other basic functions


I first started by loading the necessary packages into rstudio. 

```{r}
library(tidyverse)
library(jsonlite)
library(httr)
```


# Functions to Query Endpoints

The first function I will create will be to call data on earthquakes greater than or equal to a certain magnitude. 

```{r}
magnitude <- function(magnitude = 1) {
  a<- fromJSON('https://earthquake.usgs.gov/fdsnws/event/1/query?format=geojson')
  b<- a$features$properties %>% 
    select('mag', 'place','time','type') %>% 
    filter(mag >= magnitude)
  return(as_tibble(b))
}
```




# Contingency Tables

# Numerical Summaries Across Variables

# Plots